---
title: "Assignment 2 - Team Project"
author: "Lisa Pizzol and Blake Ellison"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2
---


```{r Section 1, message=FALSE, warning=FALSE, include=FALSE}
#Loading Libraries
library(rmarkdown)
library(bookdown)
library(knitr)
library(tidyverse)
library(readxl)
library(kableExtra)
library(plotly)

# read in data and basic tidy
population_raw <- read_excel("Data/WPP2022_GEN_F01_DEMOGRAPHIC_INDICATORS_COMPACT_REV1 (1).xlsx")
colnames(population_raw)<-population_raw[12,]
population <- population_raw[-c(1:12),] 

```


# - World and Continent Population Changes
### Data Description
The data used in this section of the assignment is from the United Nations, 2022, and can be found at [World population prospects - population division](https://population.un.org/wpp/). It contains Population, Fertility, Mortality and Migration estimates for 237 countries and/or areas from 1950 to 2021. Only the following selected variables listed in \@ref(tab:data2) have been used in Section 1 of this assignment. 

```{r data1, message=FALSE, warning=FALSE, include=FALSE}
pop1<-select(population,c(3,9,11,12,13) ) %>%
  filter(Type %in% c("World","Region"))
pop1$Year<-as.numeric(pop1$Year)
pop1$`Total Population, as of 1 January (thousands)`<-as.numeric(pop1$`Total Population, as of 1 January (thousands)`)
pop1$`Total Population, as of 1 July (thousands)`<-as.numeric(pop1$`Total Population, as of 1 July (thousands)`)

```


```{r data2, echo=TRUE, message=FALSE, warning=FALSE}
names_data<-(names(pop1))
data_type = c("Categorical","Categorical","Numerical","Numerical","Numerical")
further_information = c("Contains the following categories: World, Africa, Asia, Europe, Latin America and the Caribbean, Northern America, and Oceania.", "Contains the following information: World and Region","Contains years from 1950 to 2021", "Contains the population number in Thousands as of Janurary", "Contains the population number in Thousands as of July")
names_data_final<-cbind(names_data,data_type,further_information)
# table of variable names
kable(names_data_final,caption = "Variable Names", col.names = c('Names','Data Type','Further Information')) %>%
  kable_styling()
```

This filtered data set contains `r ncol(pop1)` variables an outline of the variable types and information is presented above in Table \@ref(tab:data2). There are `r nrow(pop1)` observations in this filtered data set. 



# References
United Nations. (2022). World population prospects - population division. https://population.un.org/wpp/ 