---
title: "Assignment 2 - Team Project"
author: "Lisa Pizzol and Blake Ellison"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2
---


```{r Section 1, message=FALSE, warning=FALSE, include=FALSE}
#Loading Libraries
library(rmarkdown)
library(bookdown)
library(knitr)
library(tidyverse)
library(readxl)
library(kableExtra)
library(plotly)

# read in data and basic tidy
population_raw <- read_excel("Data/WPP2022_GEN_F01_DEMOGRAPHIC_INDICATORS_COMPACT_REV1 (1).xlsx")
colnames(population_raw)<-population_raw[12,]
population <- population_raw[-c(1:12),] 

```


# - World and Continent Population Changes
## Data Description
The data used in this section of the assignment is from the United Nations, 2022, and can be found at [World population prospects - population division](https://population.un.org/wpp/). It contains Population, Fertility, Mortality and Migration estimates for 237 countries and/or areas from 1950 to 2021. Only the following selected variables listed in \@ref(tab:data2) have been used in this section of the assignment. 

```{r data1, message=FALSE, warning=FALSE, include=FALSE}
# select only relatvent columns and rows
pop1<-select(population,c(3,9,11,12,13) ) %>%
  filter(Type %in% c("World","Region"))
# change column type to numeric 
pop1$Year<-as.numeric(pop1$Year)
pop1$`Total Population, as of 1 January (thousands)`<-as.numeric(pop1$`Total Population, as of 1 January (thousands)`)
pop1$`Total Population, as of 1 July (thousands)`<-as.numeric(pop1$`Total Population, as of 1 July (thousands)`)

```


```{r data2, echo=TRUE, message=FALSE, warning=FALSE}
# create list of variables
names_data<-(names(pop1))
# create list of data types
data_type = c("Categorical","Categorical","Numerical","Numerical","Numerical")
# create list of detailed information
further_information = c("Contains the following categories: World, Africa, Asia, Europe, Latin America and the Caribbean, Northern America, and Oceania.", "Contains the following information: World and Region","Contains years from 1950 to 2021", "Contains the population number in Thousands as of Janurary", "Contains the population number in Thousands as of July")
#join datasets together
names_data_final<-cbind(names_data,data_type,further_information)
# table of variable information, including column names and capition for table
kable(names_data_final,caption = "Variable Information", col.names = c('Names','Data Type','Further Information')) %>%
  kable_styling()
```

This filtered data set contains `r ncol(pop1)` variables. An outline of the variable types and information is presented above in Table \@ref(tab:data2). There are `r nrow(pop1)` observations in this filtered data set. 

## Relative Changes in Population
Table \@ref(tab:data3) shows the January population (in thousands) for the entire World and the regions in 1950 and 2021. It than presents the relative population change between these two years as a percentage.

```{r data3, echo=TRUE, message=FALSE, warning=FALSE}
relative_diff<-filter(pop1,Year %in% c(1950,2021)) %>%
  select(-'Total Population, as of 1 July (thousands)',-Type) %>%
  pivot_wider(names_from = Year,values_from = 'Total Population, as of 1 January (thousands)') %>%
  mutate('Relative Difference (%)'= ((`2021`-`1950`)/`1950`)*100)

kable(relative_diff,caption = "Relative Difference of the World and Region poulations from 1950 to 2021", digits = 0) %>%
  kable_styling()
```

In Table \@ref(tab:data3) it is observed that the Worlds Relative Population Change between 1950 and 2021 is 218%, which means the population has more than doubled in the last 70 years. The table also indicates that the region with the greatest increase in population is Africa. There has been a Relative Population Change of 512% in Africa, meanwhile the region with the lowest increase is Europe with only a 36% increase. While not explored in this assignment, further research could be undertaken to unpack the wealth of people living in these regions, as this may be help gain insight to the resource support required in these regions. 

# References
United Nations. (2022). World population prospects - population division. https://population.un.org/wpp/ 